---
title: "WHO COVID-19 Data Analysis"
author: Team 3 (Xincheng Zhao;Yiheng Lu;Yu Fu;Yixiang Cheng;Qibin Gong)
output:
  html_document:
    df_print: paged
    number_sections: yes
---
```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.pos = 'H')
library(qwraps2)
library(dplyr)
library(AER)
library(ggplot2)
library(tidyverse)
options(qwraps2_markup = "markdown")
```


***

# Introduction

This project is going to investigate a dataset from WHO about COVID-19. The dataset contains number of new cases and deaths in different countries everyday since January 2020. It contains 101436 records so far and is being updated constantly. We are going to investigate this dataset to understand how the pandemic developed globally. Specifically, when did the number of new cases increases reach a maximum? Which country had the highest case-mortality rate? Does case-mortality rate differ among different regions? We are going to use EDA to answer these questions.

# Background 

Coronavirus disease 2019 (COVID-19) is caused by a new coronavirus first identified in Wuhan, China, in December 2019. Because it is a new virus, scientists are learning more each day. Although most people who have COVID-19 have mild symptoms, COVID-19 can also cause severe illness and even death. By March 05, 2021, 116 million people had COVID-19 and 2.57 million people died due to this virus. 

# Exploratory Data Analysis 

First, let's look at the daily number of new cases in the US, China and globally since Jan 3, 2020.

```{r message=FALSE, warning=FALSE, echo = FALSE}
covid <- read_csv("https://covid19.who.int/WHO-COVID-19-global-data.csv")
covid %>% group_by(Date_reported) %>%
  summarise(new_cases_global = sum(New_cases)) %>%
  ggplot(aes(y=new_cases_global, x=Date_reported)) +
  geom_line() + 
  ggtitle("Number of new cases (global)")
covid %>% group_by(Date_reported) %>% filter(Country=="United States of America") %>%
  summarise(new_cases = sum(New_cases)) %>%
  ggplot(aes(y=new_cases, x=Date_reported)) +
  geom_line() + 
  ggtitle("Number of new cases (US)")
covid %>% group_by(Date_reported) %>% filter(Country=="China") %>%
  summarise(new_cases = sum(New_cases)) %>%
  ggplot(aes(y=new_cases, x=Date_reported)) +
  geom_line() + 
  ggtitle("Number of new cases (China)")
```

We can see that the number of new cases reaches a maximum around Jan 2021 and then decreases. Recently, it decreases possibly because people are getting vaccinated. In the United States, the number of new cases has a similar trend as the global data. In China, the number of new cases exploded in February 2020, but after that it maintained at a low level. This suggests that the pandemic developed differently in different countries.

Next, let's look at the daily number of deaths in the US, China and globally since Jan 3, 2020.

```{r message=FALSE, warning=FALSE, echo = FALSE}
covid %>% group_by(Date_reported) %>%
  summarise(n_deaths_global = sum(New_deaths)) %>%
  ggplot(aes(y=n_deaths_global, x=Date_reported)) +
  geom_line() + 
  ggtitle("Number of deaths (global)")
covid %>% group_by(Date_reported) %>% filter(Country=="United States of America") %>%
  summarise(n_deaths = sum(New_deaths)) %>%
  ggplot(aes(y=n_deaths, x=Date_reported)) +
  geom_line() + 
  ggtitle("Number of deaths (US)")
covid %>% group_by(Date_reported) %>% filter(Country=="China") %>%
  summarise(n_deaths = sum(New_deaths)) %>%
  ggplot(aes(y=n_deaths, x=Date_reported)) +
  geom_line() + 
  ggtitle("Number of deaths (China)")
```

We can see that the number of deaths reaches a maximum after Jan 2021. There is about a half month lag in when it reaches its maximum compared to the number of cases. In the United States, although the number of new cases around Jan 2021 is much higher than May 2020, the number of deaths is not higher. This suggests that people are more likely to recover from COVID-19 this year. In the plot of China, there may exist an outlier around May 2020. This may be because the government did not report the number of deaths daily during a period but report the cumulative number of deaths on that day.

Case-mortality rate is an important index quantifying how dangerous a pathogen is. The global case-mortality rate so far is 2.22%. We are going to calculate the case-mortality rate so far for each country and list the top 10 countries that has the highest case-mortality rate, and top 10 countries that has the lowest case-mortality rate.

```{r message=FALSE, warning=FALSE, echo = FALSE}
covid_motality_rate <- covid %>% filter(Date_reported=="2021-03-05") %>%
  mutate(rate = Cumulative_deaths / Cumulative_cases) %>%
  select(Country, rate, WHO_region, Cumulative_cases, Cumulative_deaths)
covid_motality_rate %>%
  arrange(desc(rate)) %>%
  head(10)
covid_motality_rate %>%
  filter(rate>0) %>%
  arrange(rate) %>%
  head(10)
```

# Model

We are going to answer the question: Does case-mortality rate differ among different regions? WHO divided countries into 7 regions: Regional Office for Africa (AFRO), Regional Office for the Americas (AMRO), Regional Office for the Eastern Mediterranean (EMRO), Regional Office for Europe (EURO), Regional Office for South East Asia (SEARO), Regional Office for the Western Pacific (WPRO) and other. We are going to use an ANOVA model to compare the case-mortality rate among these regions. The model is:

$$y = \beta_0 + \beta_1AFRO+\beta_2AMRO+\beta_3EMRO+\beta_4EURO+\beta_5SEARO+\beta_6WPRO+\epsilon$$

$y$ is the case-mortality rate, and there are six dummy variables for the six regions. Other region is chosen as the baseline group. 

```{r message=FALSE, warning=FALSE, echo = FALSE}
covid_motality_rate$WHO_region <- relevel(factor(covid_motality_rate$WHO_region), ref="Other")
m <- aov(rate~WHO_region, covid_motality_rate)
summary(m)
TukeyHSD(m)
```

The ANOVA table shows that the F statistic is 2.9742 and the p-value is 0.00823. Therefore, we reject the null hypothesis and conclude that there is a difference in case-mortality rate among the 7 regions. Using Tukey's method, we can calculate a set of confidence intervals on the difference between the case-mortality rate. Based on the confidence intervals, we find that the confidence intervals for SEARO-EMRO, WPRO-EMRO and EURO-EMRO are all below 0. This means that the Eastern Mediterranean region has a significantly higher case-mortality rate than Europe, South East Asia and Western Pacific regions. 


# Sensitivity analysis

```{r message=FALSE, warning=FALSE, echo = FALSE}
plot(m, which =1:2)
covid_motality_rate[235,]
```

Based on the residual plot and QQ plot, we find that the 135th and 235th observation are outliers. Therefore, we remove it and refit the model.


```{r message=FALSE, warning=FALSE, echo = FALSE}
m <- aov(rate~WHO_region, covid_motality_rate[-c(135, 235),])
summary(m)
TukeyHSD(m)
```

After removing the outliers, the p-value in the ANOVA table is still significant. However, the Tukey's confidence intervals show that only the difference between WPRO-EMRO is significant. 

# Discussion

In Exploratory Data Analysis, we showed the trend of number of new cases and deaths since Jan 03, 2020. We found that the number of new cases and deaths reached their maximum around Jan 2021, and then decreased. This may be due to more people are getting vaccinated. We also calculate the case-mortality rate for each country, and listed the countries with the higher or lowest case-mortality rate. 

We used an ANOVA model to compare the case-mortality rate for 7 regions. After removing 2 outliers, we concluded that there is a significant difference between case-mortality rate among the 7 regions. We used Tukey's confidence interval to conclude that West Pacific region has an significant lower case-mortality rate than the Eastern Mediterranean region. This implies that Eastern Mediterranean region suffered more from the pandemic than West Pacific region and may need more help if WHO decides to distribute resources to relief the pandemic.




# Acknowledgement {-}



# Reference {-}
WHO COVID-19 Dashboard. Geneva: World Health Organization, 2020. Available online: https://covid19.who.int/



# Session info {-}
```
covid <- read_csv("https://covid19.who.int/WHO-COVID-19-global-data.csv")
covid %>% group_by(Date_reported) %>%
  summarise(new_cases_global = sum(New_cases)) %>%
  ggplot(aes(y=new_cases_global, x=Date_reported)) +
  geom_line() + 
  ggtitle("Number of new cases (global)")
covid %>% group_by(Date_reported) %>% filter(Country=="United States of America") %>%
  summarise(new_cases = sum(New_cases)) %>%
  ggplot(aes(y=new_cases, x=Date_reported)) +
  geom_line() + 
  ggtitle("Number of new cases (US)")
covid %>% group_by(Date_reported) %>% filter(Country=="China") %>%
  summarise(new_cases = sum(New_cases)) %>%
  ggplot(aes(y=new_cases, x=Date_reported)) +
  geom_line() + 
  ggtitle("Number of new cases (China)")
  
covid %>% group_by(Date_reported) %>%
  summarise(n_deaths_global = sum(New_deaths)) %>%
  ggplot(aes(y=n_deaths_global, x=Date_reported)) +
  geom_line() + 
  ggtitle("Number of deaths (global)")
covid %>% group_by(Date_reported) %>% filter(Country=="United States of America") %>%
  summarise(n_deaths = sum(New_deaths)) %>%
  ggplot(aes(y=n_deaths, x=Date_reported)) +
  geom_line() + 
  ggtitle("Number of deaths (US)")
covid %>% group_by(Date_reported) %>% filter(Country=="China") %>%
  summarise(n_deaths = sum(New_deaths)) %>%
  ggplot(aes(y=n_deaths, x=Date_reported)) +
  geom_line() + 
  ggtitle("Number of deaths (China)")

covid_motality_rate <- covid %>% filter(Date_reported=="2021-03-05") %>%
  mutate(rate = Cumulative_deaths / Cumulative_cases) %>%
  select(Country, rate, WHO_region, Cumulative_cases, Cumulative_deaths)
covid_motality_rate %>%
  arrange(desc(rate)) %>%
  head(10)
covid_motality_rate %>%
  filter(rate>0) %>%
  arrange(rate) %>%
  head(10)

covid_motality_rate$WHO_region <- relevel(factor(covid_motality_rate$WHO_region), ref="Other")
m <- aov(rate~WHO_region, covid_motality_rate)
summary(m)
TukeyHSD(m)

plot(m, which =1:2)
covid_motality_rate[235,]

m <- aov(rate~WHO_region, covid_motality_rate[-c(135, 235),])
summary(m)
TukeyHSD(m)

```



