---
title: "WHO COVID-19 Data Analysis"
author: Team 3 (Xincheng Zhao;Yiheng Lu;Yu Fu;Yixiang Cheng;Qibin Gong)
output:
  html_document:
    df_print: paged
    number_sections: yes
---
```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.pos = 'H')
library(qwraps2)
library(dplyr)
library(AER)
library(ggplot2)
library(tidyverse)
options(qwraps2_markup = "markdown")
```


***
# Abstract

COVID-19 has been circulating as a serious virus throughout the region since 2019. In this report, the number of deaths due to COVID-19 infection in different countries and regions since January 2020 will be studied. We will first briefly introduce the source of interest and purpose of the research through the "Introduction", and then expand the discussion in the "Background" section. In the section "Descriptive Analysis", we introduce relevant data in detail and develop simple information extraction. In the "Inferential Analysis" section, we will introduce polynomial regression and ANOVA modeling to analyze dataset. "Sensitivity analysis" confirmed the correctness of our model. Despite the limitations of this report, we can draw basic conclusions about the association between mortality rates and different countries based on the analysis.

# Introduction

COVID-19 is a highly contagious virus. For more than a year, almost the entire world has been affected by the virus. Analyzing and studying COVID-19 data has also become a popular trend. Outbreaks of COVID-19 have led to the deaths of many infected people. It has become a very important task to find the connection between the death toll and different countries and regions and take active countermeasures, which is also the motivation of our report. In this paper, we will investigate the WHO dataset on COVID-19. This data set contains the number of new cases and deaths that have occurred in different countries/territories on a daily basis since January 2020. So far, it contains 101,436 records and is constantly being updated. We will investigate this data set to understand how the pandemic has developed on a global scale. We will focus on: is the case fatality rate different in different regions? There are many other questions to explore, such as: When will the number of new cases peak? Which country has the highest case deaths? We will use EDA to answer these questions, and deaths in different countries will be our key variables.

# Background 

Coronavirus Disease 2019 (COVID-19) is a highly contagious respiratory disease. Since its emergence, the world has been gripped by panic. As a new virus, scientists are trying to find out more about COVID-19 every day. Although most people with COVID-19 have mild symptoms, COVID-19 can also cause severe illness and even death. By March 5, 2021, 116 million people have died from COVID-19,2.57 million have died from the virus. In this context, it is imperative to study the characteristics of COVID-19 and try to find a response. In this article, we try to use "region" as the primary entry point and use data sets from the World Health Organization (WHO). This dataset is updated daily with information on cases of COVID-19 collected in each of the 237 countries. Data collection began on January 3, 2020. In this paper, we will mainly explore: "Does the case fatality rate differ from country to country?" Wait for a problem. And based on the data we have, we can make a simple guess: "Fatality rates do vary in different countries." In other words, the mortality rate is related to the region. From the paper called "2019-novel Coronavirus (2019-NCOV): estimating the case fatality rate -- a word of caution" (2020.7), we know that "in different regions of China and different regions of the world, the sensitivity to the 2019 Novel Coronavirus is different. Immunogenetic and socioeconomic factors may contribute to differences in susceptibility to the disease."

# Descriptive Analysis 

Let's look at the daily number of new cases in the US, China and globally since Jan 3, 2020.

```{r message=FALSE, warning=FALSE}
covid <- read_csv("https://covid19.who.int/WHO-COVID-19-global-data.csv")
covid %>% group_by(Date_reported) %>%
  summarise(new_cases_global = sum(New_cases)) %>%
  ggplot(aes(y=new_cases_global, x=Date_reported)) +
  geom_line() + 
  ggtitle("Number of new cases (global)")
covid %>% group_by(Date_reported) %>% filter(Country=="United States of America") %>%
  summarise(new_cases = sum(New_cases)) %>%
  ggplot(aes(y=new_cases, x=Date_reported)) +
  geom_line() + 
  ggtitle("Number of new cases (US)")
covid %>% group_by(Date_reported) %>% filter(Country=="China") %>%
  summarise(new_cases = sum(New_cases)) %>%
  ggplot(aes(y=new_cases, x=Date_reported)) +
  geom_line() + 
  ggtitle("Number of new cases (China)")
```

We can see that the number of new cases reaches a maximum around Jan 2021 and then decreases. Recently, we are interested in what reason cause the number of new cases decreases. After we looking for information online, it's possibly because that there are several new type of covid-19 vaccines published, and people are getting vaccinated. In the United States, the number of new cases has a similar trend as the global data as most of countries globally. In China, the number of new cases exploded in February 2020, but after that it maintained at a low level, it controlled the covid-19 very well as a few other countries did. This suggests that the pandemic developed differently in different countries.

However, there has a significant data outlier for china that there has one day increased over 15000 new cases, which takes up almost half of total new numbers. After searching we find that this is because chinese government report the cases which hasn't been diagnosed earily.

Next, let's look at the daily number of deaths in the US, China and globally since Jan 3, 2020. 

```{r message=FALSE, warning=FALSE}
covid %>% group_by(Date_reported) %>%
  summarise(n_deaths_global = sum(New_deaths)) %>%
  ggplot(aes(y=n_deaths_global, x=Date_reported)) +
  geom_line() + 
  ggtitle("Number of deaths (global)")
covid %>% group_by(Date_reported) %>% filter(Country=="United States of America") %>%
  summarise(n_deaths = sum(New_deaths)) %>%
  ggplot(aes(y=n_deaths, x=Date_reported)) +
  geom_line() + 
  ggtitle("Number of deaths (US)")
covid %>% group_by(Date_reported) %>% filter(Country=="China") %>%
  summarise(n_deaths = sum(New_deaths)) %>%
  ggplot(aes(y=n_deaths, x=Date_reported)) +
  geom_line() + 
  ggtitle("Number of deaths (China)")
```

We can see that the number of deaths reaches a maximum after Jan 2021. There is about a half month lag in when it reaches its maximum compared to the number of cases. In the United States, although the number of new cases around Jan 2021 is much higher than May 2020, the number of deaths is not higher. This suggests that people are more likely to recover from COVID-19 this year. In the plot of China, there may exist an outlier around May 2020. This may be because the government did not report the number of deaths daily during a period but report the cumulative number of deaths on that day.

Case-mortality rate is an important index quantifying how dangerous a pathogen is. The global case-mortality rate so far is 2.22%. We are going to calculate the case-mortality rate so far for each country and list the top 10 countries that has the highest case-mortality rate, and top 10 countries that has the lowest case-mortality rate. An interesting fact we find is that the region for top-10 low mortality rate countries distributed equally in seven regions, but the region for top-10 high mortality rate countries are distributed more concentrated on East Mediterranean countreis and Latin America. This possible because relate to their medical service which is underdeveloped.

```{r}
covid_motality_rate <- covid %>% filter(Date_reported=="2021-03-05") %>%
  mutate(rate = Cumulative_deaths / Cumulative_cases) %>%
  select(Country, rate, WHO_region, Cumulative_cases, Cumulative_deaths)
covid_motality_rate %>%
  arrange(desc(rate)) %>%
  head(10)
covid_motality_rate %>%
  filter(rate>0) %>%
  arrange(rate) %>%
  head(10)
```

# Inferential Analysis

We are going to answer the question: Does case-mortality rate differ among different regions? WHO divided countries into 7 regions: Regional Office for Africa (AFRO), Regional Office for the Americas (AMRO), Regional Office for the Eastern Mediterranean (EMRO), Regional Office for Europe (EURO), Regional Office for South East Asia (SEARO), Regional Office for the Western Pacific (WPRO) and other. We are going to use an ANOVA model to compare the case-mortality rate among these regions. The model is:

$$y = \beta_0 + \beta_1AFRO+\beta_2AMRO+\beta_3EMRO+\beta_4EURO+\beta_5SEARO+\beta_6WPRO+\epsilon$$

$y$ is the case-mortality rate, and there are six dummy variables for the six regions. Other region is chosen as the baseline group. 

```{r}
covid_motality_rate$WHO_region <- relevel(factor(covid_motality_rate$WHO_region), ref="Other")
m <- aov(rate~WHO_region, covid_motality_rate)
summary(m)
TukeyHSD(m)
```

The ANOVA table shows that the F statistic is 2.9742 and the p-value is 0.00823. Therefore, we reject the null hypothesis and conclude that there is a difference in case-mortality rate among the 7 regions. Using Tukey's method, we can calculate a set of confidence intervals on the difference between the case-mortality rate. Based on the confidence intervals, we find that the confidence intervals for SEARO-EMRO, WPRO-EMRO and EURO-EMRO are all below 0. This means that the Eastern Mediterranean region has a significantly higher case-mortality rate than Europe, South East Asia and Western Pacific regions. 


# Sensitivity analysis

```{r message=FALSE, warning=FALSE}
plot(m, which =1:2)
covid_motality_rate[235,]
```

Based on the residual plot and QQ plot, we find that the 135th and 235th observation are outliers. Therefore, we remove it and refit the model.


```{r}
m <- aov(rate~WHO_region, covid_motality_rate[-c(135, 235),])
summary(m)
TukeyHSD(m)
```

After removing the outliers, the p-value in the ANOVA table is still significant. However, the Tukey's confidence intervals show that only the difference between WPRO-EMRO is significant. 

# Conclusion

In summary, we conducted research in this report around COVID-19, a highly infectious virus. Outbreaks of COVID-19 have led to the deaths of many infected people. We are eager to understand how the pandemic is developing on a global scale, to find links between the death toll and different countries and regions, and to adopt a positive response. Our research questions included "Do the case fatality rates differ from region to region? When did the number of new cases peak? Which country has the highest number of deaths?" It's pretty much settled. First, from descriptive analysis, we have the trend of new cases and deaths since January 3, 2020. We found that the number of new cases and deaths peaked around January 2021 and then declined. This could be because more people are being vaccinated. Subsequently, in the Inferential Analysis section, we used the ANOVA model and compared the case fatality rates in seven regions. After removing 2 outliers, we concluded that there were significant differences in case fatality rates between the 7 regions. To further investigate exactly which regions are most affected, we use Tukey's confidence interval. We found that the case fatality rate was significantly lower in the Western Pacific than in the Eastern Mediterranean. This means that the Eastern Mediterranean region will be more affected by the pandemic than the Western Pacific region and may need more help if WHO decides to allocate resources to mitigate the pandemic.
The results of the study are very important, and the practicality and constructiveness of the study are worth our consideration. Based on the analysis results and the existing authoritative literature, we can basically infer the more general conclusion that "the number of deaths is related to different countries and regions, and the case fatality rate is different in different regions". We can learn more about this finding in the future and apply it to the global fight against the pandemic. For example, we are seeing a gradual decline in the number of new cases and deaths. It is reasonable to speculate that this may be because more people are being vaccinated. If we can further verify that the vaccine has a positive effect on COVID-19 mortality, we can recommend that the vaccine be used in all regions of the world to improve the situation. In addition, we know that the western Pacific region has a significantly lower case fatality rate than the eastern Mediterranean region and is more affected by the pandemic than the western Pacific region. The reason behind it is worth further study. Different regions have different geographical environments and national conditions, which may include economic conditions, medical facilities, government policies, etc. These are constructive questions.

# References

Battegay Manuela, Kuehl Richarda, Tschudin-Sutter Saraha, Hirsch Hans H.abc, Widmer Andreas F.a, Neher Richard A.d(2020.7) 2019-Novel Coronavirus (2019-nCoV): estimating the case fatality rate – a word of caution.
https://smw.ch/article/doi/smw.2020.20203?fbclid=IwAR2SIVB53O6Of4WnRTSzfKdIsp3LEejJ3MG0QZIfFhiI-8he8xjnmx5MrNI



















